Legacy Data Confounds Genomics Studies

### 2019-07-12 LatexManuscript/legacy.tex
a word

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r  r r}
                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
{Total} & 3,550& & 11,716 & & 15,270\\
\end{tabular}
\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$. 
The grey text is the proportion of $Q$-associated variants per category.
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

@@

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

>>>>

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r  r r}
                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
{Total} & 3,550& & 11,716 & & 15,270\\
\end{tabular}
\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$ and pass the 1kGP strict mask. 
The grey text is the proportion of $Q$-associated variants per category.
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

@@

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
